---
title: "stat310_final"
author: "Rob Sisto"
date: "2023-04-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
#knitr::opts_knit$set(root.dir = "/home/risto/Nextcloud/stat_351(488)/project/data")
#setwd("/home/risto/Nextcloud/stats_310(410)/project")
```

# Start of Code - delete later

```{r import libraries}
#TODO make sure all set.seeds are 261
library(psych)
library(VGAM)
library(pROC)
library(MASS)
library(gam)
library(car)
library(knitr)
library(tidyverse)

```

```{r data input & preprocessing}
mat = read.csv("/Users/melchorronquillo/Desktop/Files/STAT310/student-mat.csv")
por = read.csv("/Users/melchorronquillo/Desktop/Files/STAT310/student-por.csv")
alcohol = rbind(mat, por)

alcohol['alc_con'] = alcohol$Walc + alcohol$Dalc
alcohol = subset(alcohol, select=-c(Walc, Dalc))
response = data.frame(alc_con=alcohol$alc_con)
  
#converting all of these numerical variables to ordinal ones
convert.to.ordinal = c('Medu', 'Fedu', 'traveltime', 'studytime', 'famrel', 'freetime', 'goout', 'health', 'failures')
for (c in convert.to.ordinal){
  alcohol[c] = factor(unlist(alcohol[c], use.names=F), order=T)
}  


#converting all these variables to factors
convert.to.factor = c("romantic", "internet", "higher", "nursery", "activities", "paid", "famsup", "schoolsup", "guardian", "reason", "Fjob", "Mjob", "Pstatus", "famsize", "address", "sex", "school")
for (c in convert.to.factor){
  alcohol[c] = factor(unlist(alcohol[c], use.names=F))
}  

old.age = data.frame(age=alcohol$age)
alcohol$age[alcohol['age'] >= 20] = 20 #changing all ages above 20 to 20

#key: 1 = 0=0; 1=1:2, 2=3:6, 3= >6
#binning the absences variable 
old.absences = data.frame(absences= alcohol$absences)
alcohol$absences = alcohol%>% dplyr::select(absences) %>% mutate(ab = cut(absences, breaks=c(0,0.1, 2,6, 75), include.lowest = T, labels=c(0,1,2,3))) %>% dplyr::select(ab)
alcohol$absences = factor(unlist(alcohol$absences, use.names=F), order=T)

#converting the response to tertiary for now
alcohol$consump = cut(alcohol$alc_con,
                  breaks=c(0,3,5,10),
                  labels=c("low", "medium", "high")
)
alcohol$consump = factor(unlist(alcohol$consump, use.names=F), order=T)
alcohol = subset(alcohol, select=-alc_con)
```

```{r splitting the data into test/train for tertiary response}
set.seed(261)
sample <- sample(c(TRUE, FALSE), nrow(alcohol), replace=TRUE, prob=c(0.7,0.3))
train =alcohol[sample,]
test = alcohol[!sample,]
```

```{r cumulative logit}
ord = vglm(consump~., family = cumulative(parallel = T), data = train)

#choosing the highest probability category for each obs
pred = predict(ord, test, type="response") 
resp = c() 
for (p in 1:nrow(pred)) {
  p = pred[p,]
  r = rank(p)
  ans = ifelse(r["low"] == 3, "low", ifelse(r["medium"] == 3, "medium", "high"))
  resp = append(resp, ans)
}

pred.ord = factor(unlist(resp, use.names=F), order=T, levels=c("low", "medium", "high"))
accu.ord = mean(pred.ord == test$consump)


#stepAIC for vglms
step.ord = step4vglm(ord)

pred = predict(step.ord, test, type="response") 
resp = c()
for (p in 1:nrow(pred)) {
  p = pred[p,]
  r = rank(p)
  ans = ifelse(r["low"] == 3, "low", ifelse(r["medium"] == 3, "medium", "high"))
  resp = append(resp, ans)
}
pred.stepord = factor(unlist(resp, use.names=F), order=T, levels=c("low", "medium", "high"))
accu.stepord = mean(pred.stepord == test$consump)


#auc for stepAIC is less than regular model
rocplot.ord = multiclass.roc(consump~fitted(ord), data=train, levels=c("low", "medium", "high"))
auc_mod.ord = auc(rocplot.ord)


#auc for stepAIC is less than regular model
rocplot.stepord = multiclass.roc(consump~fitted(step.ord), data=train, levels=c("low", "medium", "high"))
auc_mod.stepord = auc(rocplot.stepord)

```

```{r preprocessing for binary response}
#recoding the reponse variable into a binary variable
ter.response = data.frame(consump=alcohol$consump) 
alcohol$consump = cut(response$alc_con,
                  breaks=c(0,3,10),
                  labels=c("low", "high")
)
set.seed(261)
sample <- sample(c(TRUE, FALSE), nrow(alcohol), replace=TRUE, prob=c(0.7,0.3))
train =alcohol[sample,]
test = alcohol[!sample,]
```

```{r glm for binary response}
#creating logistic model
mod.glm = glm(consump~., family=binomial, data=train)

rocplot.glm = roc(consump~fitted(mod.glm), data=train)
auc_mod.glm = auc(rocplot.glm)

#predicting the responces of the test data using the inital logit model
pred = predict(mod.glm, test, type="r") 
pred.glm = ifelse(pred<0.5, "low", "high")
accu.glm = mean(pred.glm == test$consump)


#creating new logistic model using stepAIC for variable selection
step.glm = stepAIC(mod.glm, direction="backward")

rocplot.stepglm = roc(consump~fitted(step.glm), data=train)
auc_mod.stepglm = auc(rocplot.stepglm)

#predicting the responses of test data using the stepAIC model
pred = predict(step.glm, test, type="r") 
pred.stepglm = ifelse(pred<0.5, "low", "high")
accu.stepglm = mean(pred.stepglm == test$consump)

```

```{r probit model for binary response}
#creating inital probit model
mod.probit = glm(consump~., family=binomial(link="probit"), data=train)

rocplot.probit = roc(consump~fitted(mod.probit), data=train)
auc_mod.probit = auc(rocplot.probit)

#predicting test responses using initial probit
pred = predict(mod.probit, test, type="r") 
pred.probit = ifelse(pred<0.5, "low", "high")
accu.probit = mean(pred.probit == test$consump)


#creating probit model using stepAIC
step.probit = stepAIC(mod.probit, direction="backward")

rocplot.stepprobit = roc(consump~fitted(step.probit), data=train)
auc_mod.stepprobit = auc(rocplot.stepprobit)

#predicting responses using improved probit model
pred = predict(step.probit, test, type="r") 
pred.stepprobit = ifelse(pred<0.5, "low", "high")
accu.stepprobit = mean(pred.stepprobit == test$consump)

```

```{r gam for binary response}
#creating a gam model
mod.gam = gam::gam(consump~., family=binomial, data=train)

rocplot.gam = roc(consump~fitted(mod.gam), data=train)
auc_mod.gam = auc(rocplot.gam)

#predicting responses for gam
pred = predict(mod.gam, test, type="r") 
pred.gam = ifelse(pred<0.5, "low", "high")
accu.gam = mean(pred.gam == test$consump)


#creating a gam model using only significant variables
new.gam = gam::gam(consump~goout+health+absences+famrel+studytime+Fjob+Fedu+Medu+famsize+age+sex, family=binomial, data=train)

rocplot.newgam = roc(consump~fitted(new.gam), data=train)
auc_mod.newgam = auc(rocplot.gam)


#predicting responses using improved gam model
pred = predict(new.gam, test, type="r") 
pred.newgam = ifelse(pred<0.5, "low", "high")
accu.newgam = mean(pred.newgam == test$consump)


#not working
#step.Gam(mod.gam, gam.scope(alcohol, response=32))
#s.mod = stepAIC(g.mod, method="both")

```

################################################################# 

# Introduction - start of report

# Appendix A

```{r appendix}
#beginning of Appendix Code
knitr::opts_chunk$set(include = T)
knitr::opts_chunk$set(echo = F)

prez.alcohol = alcohol
prez.alcohol$consump = response$alc_con

#TODO decide if we should place the "Figure X" header before or after the figure itself
```

### Figure 1: EDA on Pre-Binned Absences Variable
```{r figure 1}
#EDA on Pre-Binned Absences Variable

ggplot(old.absences, aes(absences)) + 
  geom_histogram(binwidth=1) + ggtitle("Distribution of Pre-Binned Absences Variable")

ggplot(old.absences, aes(y=absences)) + 
  geom_boxplot() + ggtitle("Boxplot of Pre-Binned Absences Variable")

kable(describe(as.numeric(unlist(old.absences$absences)), quant=c(0.25,0.75)), caption="Descriptive Statistics for Pre-Binned Absences Variable")
```

### Figure 2: EDA on the Age Variable (Binned)
```{r figure 2}
#eda on the Age Variable
ggplot(old.age, aes(age)) + 
  geom_histogram(binwidth = 0.5) + ggtitle("Distribution of Pre-Binned Age Variable")

kable(prop.table(table(Age = prez.alcohol$age)), caption="Proportion of Student Age after Binning") #proportion of each student age after it was binned 
#TODO maybe dont include this proportion table 
```


###Figure 3:Pair Plots of variables with WEAKER than expected correlation
```{r figure 3}
pairs(prez.alcohol[c('consump', 'age', 'Medu', 'Fedu', 'famrel', 'health')], pch=19, col='orange', lower.panel=panel.smooth)
```

###Figure 4:Pair Plots of variables with STRONGER than expected correlation
```{r figure 4}
pairs(prez.alcohol[c('consump', 'goout', 'freetime', 'sex', 'studytime')], pch=19, col='orange', lower.panel=panel.smooth)
```

### Figure 5: Relationship between Alcohol Consumption and Goout Variable
```{r figure 5}
#eda on Relationship between Alcohol Consumption and Goout Variable
mosaicplot(table(prez.alcohol$goout, prez.alcohol$consump), shade=T, main="Alcohol Consumption and Going Out")
```

### Figure 6: Relationship between Alcohol Consumption and Freetime Variable
```{r figure 6}
#eda on Relationship between Alcohol Consumption and Freetime Variable
mosaicplot(table(prez.alcohol$freetime, prez.alcohol$consump), shade=T, main="Alcohol Consumption and Free time")
```

### Figure 7: Relationship between Alcohol Consumption and Studytime Variable
```{r figure 7}
#eda on Relationship between Alcohol Consumption and Freetime Variable
mosaicplot(table(prez.alcohol$studytime, prez.alcohol$consump), shade=T, main="Alcohol Consumption and Study time")
```

### Figure 8: Relationship between Ordinal Alcohol Consumption by Sex
```{r figure 8}
#eda on Relationship between Ordinal Alcohol Consumption by Sex
mosaicplot(table(prez.alcohol$sex, prez.alcohol$consump), shade=T, main="Alcohol Consumption and Study time")
```


### ****Figure 9: Relationship between Ordinal Alcohol Consumption by Age
```{r figure 9}
#eda on Relationship between Ordinal Alcohol Consumption by Age
cont.age = table(prez.alcohol$age, prez.alcohol$consump)
mosaicplot(cont.age, shade=T, main='Mosaic Plot of Age and Alcohol Consumption')

ggplot(prez.alcohol, aes(x=factor(consump), y=age)) + 
  geom_boxplot() + ggtitle("Boxplot of Alcohol Consumption by Age") + xlab("Alcohol Consumption") + ylab("Age")
```


### Figure 9: EDA on Ordinal Alcohol Consumption
```{r figure 9}
#eda on ordinal alcohol consumption
kable(describe(response$alc_con, quant=c(0.25,0.75)), caption="Descriptive Statistics for Ordinal Alcohol Consumption")

ggplot(response, aes(alc_con)) + 
  geom_histogram(binwidth = 0.5) + ggtitle("Distribution of Ordinal Alcohol Consumption") + xlab("Alcohol Consumption")
```

### Figure 10: EDA on Tertiary Alcohol Consumption
```{r figure 10}
#eda on Tertiary Alcohol Consumption
kable(describe(as.numeric(unlist(ter.response$consump)), quant=c(0.25,0.75)), caption="Descriptive Statistics for Tertiary Alcohol Consumption")

ggplot(ter.response, aes(as.numeric(consump))) + 
  geom_histogram(binwidth = 0.5) + ggtitle("Distribution of Tertiary Alcohol Consumption") + xlab("Alcohol Consumption")
```

### Figure 11: EDA on Binary Alcohol Consumption
```{r figure 11}
#eda on binary alcohol consumption
ggplot(alcohol, aes(as.numeric(consump))) + 
  geom_histogram(binwidth = 0.5) + ggtitle("Distribution of Binary Alcohol Consumption") + xlab("Alcohol Consumption")

```

# EDA - delete this
```{r eda}
#TODO delete this code chunk once everything has been organized into separate figures

#TODO add main pairs plot that melchor made of all the important variables from the presentation

#TODO add any other plots/graphs/tables that melchor thinks are important to add or that we want to talk about in our report

```

### Figure X: Initial Cumulative Logit Model Metrics
```{r figure x}
#TODO rename this figure

#for original model
a = Anova(ord)
a = as.data.frame(a)
a$Sig = ifelse(a$`Pr(>Chisq)`<= 0.001, "***", ifelse(a$`Pr(>Chisq)`<= 0.01, "**", ifelse(a$`Pr(>Chisq)`<= 0.05, "*", ifelse(a$`Pr(>Chisq)`<= 0.1, ".", " "))))

kable(a, caption="Analysis of Deviance Table (Type II tests)")


summary(ord) #TODO see if we can find out another way to print this more prettily, otherwise its fine as is

kable(addmargins(table(predicted = pred.ord, actual = test$consump)), caption="Contingency Table of Predicted vs. Actual Values using Inital Cumulative Logit Model")


kable(data.frame("Accuracy on Test Data" = accu.ord, "AUC from ROC Curve"= auc_mod.ord), caption="Various Metrics from the Inital Cumulative Logit Model")

#plotting not working for the multiclass.rocplot method
#plot(rocplot.ord)

```

### Figure X: Improved Cumulative Logit Model Metrics
```{r figure x}
#TODO rename figure

#for the stepAIC model
a = Anova(step.ord)
a = as.data.frame(a)
a$Sig = ifelse(a$`Pr(>Chisq)`<= 0.001, "***", ifelse(a$`Pr(>Chisq)`<= 0.01, "**", ifelse(a$`Pr(>Chisq)`<= 0.05, "*", ifelse(a$`Pr(>Chisq)`<= 0.1, ".", " "))))
kable(a, caption="Analysis of Deviance Table (Type II tests)")

summary(step.ord)


kable(addmargins(table(predicted = pred.stepord, actual = test$consump)), caption="Contingency Table of Predicted vs. Actual Values using Improved Cumulative Logit Model")

kable(data.frame("Accuracy on Test Data" = accu.stepord, "AUC from ROC Curve"= auc_mod.stepord), caption="Various Metrics from the Improved Cumulative Logit Model")

#plot(rocplot.stepord)

```


### Figure X: Model Metrics for Inital GLM
```{r figure x}
#TODO rename figure
#for original glm

a = Anova(mod.glm)
a = as.data.frame(a)
a$Sig = ifelse(a$`Pr(>Chisq)`<= 0.001, "***", ifelse(a$`Pr(>Chisq)`<= 0.01, "**", ifelse(a$`Pr(>Chisq)`<= 0.05, "*", ifelse(a$`Pr(>Chisq)`<= 0.1, ".", " "))))
kable(a, caption="Analysis of Deviance Table (Type II tests)")


summary(mod.glm) #TODO maybe change this to a more pretty knitted output


plot(rocplot.glm, main="Plot of the ROC Curve for the Initial GLM")

kable(addmargins(table(predicted = pred.glm, actual = test$consump)), caption="Contingency Table of Predicted vs. Actual Values using Initial Logit Model")

kable(data.frame("Accuracy on Test Data" = accu.glm, "AUC from ROC Curve"= auc_mod.glm), caption="Various Metrics from the Inital Logit Model")

```



### Figure X: Model Metrics for Improved GLM
```{r}
#TODO rename figure
#for stepAIC glm

#TODO the output for this looks weird, try to make it look better
a = Anova(step.glm)
a = as.data.frame(a)
a$Sig = ifelse(a$`Pr(>Chisq)`<= 0.001, "***", ifelse(a$`Pr(>Chisq)`<= 0.01, "**", ifelse(a$`Pr(>Chisq)`<= 0.05, "*", ifelse(a$`Pr(>Chisq)`<= 0.1, ".", " "))))
kable(a, caption="Analysis of Deviance Table (Type II tests)")

summary(step.glm)

plot(rocplot.stepglm, main="Plot of the ROC Curve for the Improved GLM")

kable(addmargins(table(predicted = pred.stepglm, actual = test$consump)), caption="Contingency Table of Predicted vs. Actual Values using the Improved Logit Model")

kable(data.frame("Accuracy on Test Data" = accu.stepglm, "AUC from ROC Curve"= auc_mod.stepglm), caption="Various Metrics from the Improved Logit Model")

```


### Figure X: Model Metrics for Initial Probit
```{r}
#TODO rename figure

#original probit model
a = Anova(mod.probit)
a = as.data.frame(a)
a$Sig = ifelse(a$`Pr(>Chisq)`<= 0.001, "***", ifelse(a$`Pr(>Chisq)`<= 0.01, "**", ifelse(a$`Pr(>Chisq)`<= 0.05, "*", ifelse(a$`Pr(>Chisq)`<= 0.1, ".", " "))))
kable(a, caption="Analysis of Deviance Table (Type II tests)")

summary(mod.probit)

plot(rocplot.probit, main="Plot of the ROC Curve for the Inital Probit")

kable(addmargins(table(predicted = pred.probit, actual = test$consump)), caption="Contingency Table of Predicted vs. Actual Values using the Inital Probit Model")

kable(data.frame("Accuracy on Test Data" = accu.probit, "AUC from ROC Curve"=auc_mod.probit), caption="Various Metrics from the Inital Probit Model")

```


### Figure X: Model Metrics for the Improved Probit
```{r}
#TODO rename figure

#stepAIC probit model

a = Anova(step.probit)
a = as.data.frame(a)
a$Sig = ifelse(a$`Pr(>Chisq)`<= 0.001, "***", ifelse(a$`Pr(>Chisq)`<= 0.01, "**", ifelse(a$`Pr(>Chisq)`<= 0.05, "*", ifelse(a$`Pr(>Chisq)`<= 0.1, ".", " "))))
kable(a, caption="Analysis of Deviance Table (Type II tests)")

summary(step.probit)

plot(rocplot.stepprobit, main="Plot of the ROC Curve for the Improved Probit")

kable(addmargins(table(predicted = pred.stepprobit, actual = test$consump)), caption="Contingency Table of Predicted vs. Actual Values using the Improved Probit Model")

kable(data.frame("Accuracy on Test Data" = accu.stepprobit, "AUC from ROC Curve"=auc_mod.stepprobit), caption="Various Metrics from the Improved Probit Model")

```


### Figure X: Model Metrics for the Inital GAM
```{r}
#TODO rename figure
summary(mod.gam)
plot(rocplot.gam, main="Plot of the ROC Curve for the Initial GAM")

kable(addmargins(table(predicted = pred.gam, actual = test$consump)), caption="Contingency Table of Predicted vs. Actual Values using the Initial GAM")

kable(data.frame("Accuracy on Test Data" = accu.gam, "AUC from ROC Curve"=auc_mod.gam), caption="Various Metrics from the Initial GAM")

```


### Figure X: Model Metrics for the Improved GAM
```{r}
summary(new.gam)
plot(rocplot.newgam, main="Plot of the ROC Curve for the Improved GAM")


kable(addmargins(table(predicted = pred.newgam, actual = test$consump)), caption="Contingency Table of Predicted vs. Actual Values using the Improved GAM")

kable(data.frame("Accuracy on Test Data" = accu.newgam, "AUC from ROC Curve"=auc_mod.newgam), caption="Various Metrics from the Improved GAM")
```

### Figure X: Comparing Model Metrics
```{r}
metrics = data.frame("Model" = c("Initial Cummulative Logit", "Improved Cumulative Logit", "Initial GLM", "Improved GLM", "Inital Probit Model",
                                 "Improved Probit Model", "Inital GAM", "Improved GAM"),
                     "AUC" = c(auc_mod.ord, auc_mod.stepord, auc_mod.glm, auc_mod.stepglm, auc_mod.probit, auc_mod.stepprobit, auc_mod.gam, auc_mod.newgam), 
                     "Accuracy on Test Data" = c(accu.ord, accu.stepord, accu.glm, accu.stepglm, accu.probit, accu.stepprobit, accu.gam, accu.newgam))

kable(metrics)
```


# Appendix B

### Code

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
#this chunk prints out all of the code
```
